---
title: "check-mhap-target-snp-MAF-assignments"
author: "Neil Thompson"
date: "2024-04-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## check to see if microhaplotype assignments match the target SNP assignment

```{r}
library(tidyverse)
library(knitr)
library(kableExtra)
library(readxl)
```

#load assignments 
```{r, load-assignments}
mhap <- readRDS("../data-microtyper/USDA-ckmr-microhap-PO-assignments.rds") %>%
  dplyr::select(-"relationship_class") %>%
  pivot_longer(sire:dam, names_to="sex", values_to = "brood_id") %>%
  mutate(marker = "microhaplotype") %>%
  filter(!is.na(brood_id))
  
snp <- readRDS("../data-ckmr/ckmr-po-hotspot-USDA.rds") %>%
  mutate(marker = "snp_target")

mhap_logls <- readRDS("../data-microtyper/USDA-ckmr-microhap-logl-values-allpairwisecomps.rds") %>%
  rename(mhap_logl = logl_ratio,
            mhap_num_loc = num_loc)
```

How many juveniles are in MH assignments?
```{r}
mhap %>% filter(!is.na(brood_id)) %>%distinct(kid) #50 kids, 2 kids didnt pass QC
mhap_logls%>%distinct(D2_indiv) #53 kids in the full MH data set
50/53 #microhap assignment rate 
mhap %>% count(kid) %>%ungroup() %>% count(n) #trios
```

did the top two SNP assignments match the microhap assignments?
```{r}
snp %>% group_by(D2_indiv) %>%
  slice(1:2) %>%
  ungroup() %>%
  left_join(mhap, . , by =c("kid"="D2_indiv", "brood_id" = "D1_indiv")) %>% #join mh assignments to snp assignments asking did the snp get the same answer as the mhap?
  arrange(kid) %>%
  filter(is.na(logl_ratio))
```


why are these discordant?

G005404
```{r}
snp %>% filter(D2_indiv=="G005404", D2_indiv=="G001785") 
```
Missing parent not included in the SNP analysis.

G005504
```{r}
snp%>% filter(D2_indiv=="G005504") %>% slice(1:5) %>%
  left_join(., mhap_logls, join_by("D2_indiv", D1_indiv))
```
True parent ranked 3rd in target SNP analysis

G005519
```{r}
snp %>% filter(D2_indiv=="G005519", D2_indiv=="G005462")  
```
Missing parent not included in the SNP analysis.

G005527
```{r}
snp%>% filter(D2_indiv=="G005527") %>% slice(1:5) %>%
  left_join(., mhap_logls, join_by("D2_indiv", D1_indiv))
```

True parent ranked 3rd in target SNP analysis

How many kids assigned using target SNP Marker?
```{r}
snp_assigns <- snp %>% group_by(D2_indiv) %>%
  slice(1:2) %>%
  ungroup() %>%
  filter(logl_ratio>=10) %>% #logl_ratio at left end of PO logl distribution plot
  distinct(D2_indiv, .keep_all = T) %>%
  mutate(mh_kid = D2_indiv%in%mhap$kid)

snp_assigns %>% 
  filter(!D2_indiv=="G005417") %>% # removing because brood and kid spawned in different years
  left_join(., mhap, by=c("D2_indiv"="kid", "D1_indiv"="brood_id")) 

snp_kids <- snp_assigns %>% distinct(D2_indiv) %>% filter(!D2_indiv=="G005417") %>% pull(D2_indiv)
mhap %>% filter(!kid%in%snp_kids) #G005519 not in data

snp %>% filter(D2_indiv=="G005519") %>% filter(D1_indiv%in% c("G005462","G005463"))
```
G005417 and G005418 relationship removed (as it was in MH) because that kid and
parent were spawned in different years.

how many kids in SNP dataset?
```{r}
snp %>% distinct(D2_indiv) %>% nrow(.)#53
```
how many trios?
```{r}
 snp %>% 
  group_by(D2_indiv) %>%
  slice(1:2) %>%
  ungroup() %>%
  filter(logl_ratio>=10) %>%group_by(D2_indiv) %>% count(D2_indiv) %>% 
  ungroup() %>%count(n) # 2 is trio, 1 is duo relationship.
```

#MAF assignment comparisons

load in MAF assignments
```{r}
maf <- readRDS("../data-ckmr/ckmr-po-mafSNP-USDA.rds") 
head(maf) #all assignments from MAF

maf_assigns <- maf %>% group_by(D2_indiv) %>%
  arrange(desc(logl_ratio)) %>%
  slice(1:2) %>%
  ungroup() %>%
  filter(logl_ratio>=10) %>% #logl_ratio at left end of PO logl distribution plot
  distinct(D2_indiv, .keep_all = T) %>%
  mutate(mh_kid = D2_indiv%in%mhap$kid)

nrow(maf_assigns) # number of kids assigned 
```

did MAF assigns match MH assigns?
```{r}
maf %>%
  group_by(D2_indiv)  %>%
  arrange(desc(logl_ratio)) %>%
  slice(1:2) %>%
  ungroup() %>%
  left_join(mhap, . , by =c("kid"="D2_indiv", "brood_id" = "D1_indiv")) %>% #join mh assignments to snp assignments asking did the snp get the same answer as the mhap?
  arrange(kid) %>%
  filter(is.na(logl_ratio))
```
2 kids didn't match the MH assignments. Assigned in MH, but not MAF
did the kids assign in MAF?
```{r}
maf_assigns %>% filter(D2_indiv%in%c("G005405", "G005409"))
```
yep, kids assigned, but they were single parent assignments.
```{r}
mhap %>% filter(kid%in%c("G005405", "G005409")) %>%
  left_join(., mhap_logls, by=c("kid"="D2_indiv", "brood_id"="D1_indiv"))
```
Both parents assigned with plenty large logl ratios in microhaps.

Now, why only 46 kids assign in the MAF? that seems low.

```{r}
maf %>%distinct(D2_indiv) %>%nrow(.) #53 spat in the analysis, same as SNP and MH
maf_kids <- maf_assigns %>% distinct(D2_indiv) %>% pull(D2_indiv)

maf %>% filter(!D2_indiv%in%maf_kids) %>% group_by(D2_indiv) %>% slice(1:5)
```
G005407 low log single parent assign MH
G005412 no assign --> same in MH
G005414 no assign --> same in MH
G005417 no assign --> same in MH --NT note unassigned because the only related 
parent was used in a different year class. The broodstock assigned had a high 
number of loci (n=69) with 3 or more alleles passing QC filtering in the MH 
data. Suggests issue with library preparation or genotyping quality.
G005512 no assign low logs correct parents
G005515 low logs correct parents
G005534 low log correct parent. (duo in MH)

All the relationships are there, it's just that MAF didn't pass logl filtering
threshold of 10.

how many trios in MAF?
```{r}
maf %>% group_by(D2_indiv) %>%
  arrange(desc(logl_ratio)) %>%
  slice(1:2) %>%
  ungroup() %>%
  filter(logl_ratio>=10) %>%group_by(D2_indiv) %>% count(D2_indiv) %>% 
  ungroup() %>%count(n) # 2 is trio, 1 is duo relationship.
```


Make a BIG table with the mhap assignments as truth, join the SNP and MAF to it.
```{r}
snp2join <- snp %>% 
  group_by(D2_indiv) %>%
  arrange(desc(logl_ratio)) %>%
  slice(1:2) %>%
  ungroup()

mh2join <- left_join(mhap,mhap_logls, by=c("kid"="D2_indiv","brood_id"="D1_indiv")) 

all_assigns <- maf %>% group_by(D2_indiv) %>%
  arrange(desc(logl_ratio)) %>%
  slice(1:2) %>%
  ungroup() %>%
  mutate(mtype = "MAF") %>%
  left_join(., snp2join, by=c("D2_indiv"="D2_indiv", "D1_indiv"="D1_indiv")) %>%
  rename(maf_logl = logl_ratio.x,
         maf_nloc = num_loc.x,
         snp_logl = logl_ratio.y,
         snp_nloc = num_loc.y) %>%
  left_join(mh2join,.,by=c("kid" = "D2_indiv", "brood_id"= "D1_indiv")) #whole enchilada to summarize

all_assigns
```

From Tom: I think a discordant assignments paragraph in the results would be helpful. It could address discordances between the different methods and the pedigree, as you do with the two cases in Table 1.

(NT in Power for relationship reconstruction); Another thing to consider (I’m not insisting, just an idea for consideration) might be to take all the offspring with assignments that are concordant across panels and with the pedigree (all panels and the pedigree agree, so we are pretty confident and it is a common set to use for comparison) and compare the mean difference in LLR between the true parent-offspring pairs and the unrelated pairs. It could be a numerical way to reinforce the point you make visually in Figures 1 and 2.

Calc Tom's statistic:

```{r}
meta_po <- read_xlsx("../data/thermofisher-sutherland-allsamples_metadata.xlsx", 
                     sheet = "metadata_PO")
ped_long <- meta_po %>% dplyr::select(-fam) %>%
  pivot_longer(-indiv_id, names_to="sex", values_to = "brood_id") %>% 
  dplyr::select(-sex)

#join Mhap logls to it
ped_long %>%
  left_join(., mhap_logls, by=c("indiv_id"="D2_indiv", "brood_id"="D1_indiv")) %>%
  left_join(., snp, by=c("indiv_id"="D2_indiv", "brood_id"="D1_indiv")) %>%
  rename(snp_logl = logl_ratio, 
         snp_num_loc = num_loc) %>% dplyr::select(-marker) %>%
  left_join(., maf,  by=c("indiv_id"="D2_indiv", "brood_id"="D1_indiv")) %>%
  rename(maf_logl = logl_ratio, 
         maf_num_loc = num_loc) %>%
  filter(!(is.na(mhap_logl)|is.na(snp_logl)|is.na(maf_logl))) %>%
  filter(indiv_id %in% mhap$kid)
```

which assignments match pedigree?
```{r}
maf2join <- maf %>% filter(D2_indiv %in% maf_assigns$D2_indiv) %>%
  filter(logl_ratio>=10) %>% group_by(D2_indiv) %>% slice(1:2) %>% ungroup()

inner_join(mhap, ped_long, join_by("kid"=="indiv_id", "brood_id"=="brood_id"))  %>% #mhap to ped
  inner_join(., snp2join%>%filter(logl_ratio>=10), join_by("kid"=="D2_indiv", "brood_id"=="D1_indiv")) %>% #SNP to mhap to ped 
  inner_join(., maf2join, join_by("kid"=="D2_indiv", "brood_id"=="D1_indiv")) %>%
  left_join(., mhap_logls, join_by("kid"=="D2_indiv", "brood_id"=="D1_indiv")) %>%
  rename(snp_logl = logl_ratio.x,
         snp_loc = num_loc.x,
         maf_logl = logl_ratio.y,
         maf_loc = num_loc.y) %>%
  group_by(kid) %>%
  summarise(
    rel_type = case_when(
    n()==1 ~"duo",
    n()==2 ~"trio"),
    mhap_loc = mean(mhap_num_loc),
    mhap_logl = mean(mhap_logl),
    snp_loc = mean(snp_loc),
    snp_logl = mean(snp_logl),
    maf_loc = mean(maf_loc),
    maf_logl = mean(maf_logl)) %>%
  mutate(mhap_snp = mhap_logl-snp_logl,
         mhap_maf = mhap_logl-maf_logl,
         ln_mh_snp = mhap_loc-snp_loc,
         ln_mh_maf = mhap_loc-maf_loc) %>%
  dplyr::select(kid, rel_type, mhap_snp, ln_mh_snp, mhap_maf, ln_mh_maf) %>%
  mutate(mh_snp_mean = mean(mhap_snp),
         mh_maf_mean = mean(mhap_maf))
```

#le fin.
